---
author: "Kamal Sacranie"
numbersections: true
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: bookdown::pdf_document2
---

# (PART) Week 1 {-}

```{r, include=F}
knitr::opts_chunk$set(echo = TRUE)
```

# Simple plotting with `ggplot`

What is the association between GDP and life-expectancy:

```{r, echo=FALSE}
library(gapminder)
library(tidyverse)
```

## Using `ggplot`

::: {.example #gdp-life-exp name="GDP vs life expectancy"}
<br />\hfill\break

::: {.minipage latex-data=""}

```{r ggplottest, fig.cap="The cars data."}
#| fig.pos = "!H",
ggplot(
  # Piping to mutate to boolean index the data
  gapminder %>% mutate(
    # Simple boolean logic here and passing it as a kwarg to the mutate
    # function
    indicator = (country == "United Kingdom")
  ),
  aes(x = gdpPercap, y = lifeExp)
) +
  # How the fuck has indicator now got local scope. Was it created by the
  # mutate fucntion. But even then? it's not returning anything, where is
  # idnicator being assigned
  geom_point(aes(color = indicator))
```

:::

:::

---

- `aes` stands for aesthetic and creates an aesthetic mapping between our
  dataset and the aesthetic properties of the plot. It's a function which is
  passed as an in place arg called a `mapping` function
- `aes` takes an x variable and y variable form our dataset as arguments
- `geom_point` is one of the many plots that come with `ggplot` and plots a
  point plot

> The `aes` argument which goes in the first `ggplot` function applier for the
> whole "canvas", whereas the ones for the secondary functions only apply to
> that specific function. So **if you want to apply colour to only the
> regression line, you do the `aes` in the `geom_smooth` function**

In the above example we use the `%>%` operator (provided by the package dyplr).

This is maybe the only good thing about R that I have encoutered. It works
similarly to the pipe character in bash. It takes the return from the last
function pipes it into the first argument of the next function.

> For the record, it is insanely retarded that we can just pass through the
> columns of `gapminder` as if they were variables. We haven't assigned them or
> anything. It's also insane to me that we use ggplot and then ADD A PLOT?? WTF
> does that even mean. Why are these two objects able to be added. How is there
> not a type error

## The pipe `%>%`

As previously mentioned... the only good thing about `R`. Just plugs the data
into the first argument of the next function.

::: {.example #pipe-vs-normal name="Piping to a function"}
<br />\hfill\break

```{r}
gapminder %>% nrow()
```

is the same as

```{r}
nrow(gapminder)
```

:::

---

This is often what `R` uses to stack things. It's basically becasue `R` is made
by people who are too retarded to read regular nested code or even just regular
lines of code.

## Filtering for a value in column (basically boolean indexing)

```{r}
# notice that we can pass a straight int here. R knows the type in the column
base_plot <- ggplot(
  # Easier to see here that mutate basically adds a column with a boolean value
  # showing whether it's the UK or not
  filter(gapminder, year == 2007) %>%
    mutate(indicator = (country == "United Kingdom")),
  aes(x = gdpPercap, y = lifeExp)
) + # The plus is similar to the pipe, it is introduces by ggplot and basically
  # allows you to build layers on your plot
  geom_point(aes(color = indicator))
```

- The `+` is like ggplot's version of the pipe. It's used to layer graphs
- `indicator` is accessable because it's a column of our `gapminder` data which
  is being pushed through with our plus

```{r}
base_plot +
  geom_smooth()
```

- The line curves around the points and fits locally to the points
- The gray area is the confidence intreval of the fit line

## Model objects

Everything is an object in `R` and our assignment operator is `<-`:

```{r}
gm_2007 <- gapminder %>%
  filter(year == 2007)
```

Great so we can use that everywhere in this `R` file. For example, we can use
the `lm` function for fitting models

::: {.example #lm-basic-fit name="A basic model function use"}
<br />\hfill\break

```{r}
gm_lm <- lm(lifeExp ~ gdpPercap, data = gm_2007)
```

:::

---

> A common way for model funciton to work in `R` is to require the assignment
> of the lefthand side and the right hand side variables; we see this in
> \@ref(exm:lm-basic-fit) with the use of the `~`. On the left of the tilde we
> have the dependent (outcome) variable and on the right we have the
> independent (predictor) variable

In \@ref(exm:lm-basic-fit) `lm` stands for linear model (a regression) and is
an object which contains all the information for our regression. When you print
it on its ones, there isn't much you can do

```{r}
summary(gm_lm)
residuals(gm_lm) # the difference in the predicted values and the actual
predict(gm_lm)
```

- `summary` is a builtin function which displays a summary **of any model
  object in `R`**
- This also shows us standard errors, p-values, etc.

> Something I find weird here is that all the data we need is inside the `lm`
> object but instead of using a method to produce the result like I would
> expect, we have to pass it to another function

## `broom`

`broom` is just a library that provides nice out of the box rendering for `R`'s
model objects.

```{r}
library(broom)
```

```{r}
tidy(gm_lm)
```

- The `broom::tidy` function is really nice for making tables look good
- Don't think it works in latex though

We also have the `broom::glance` function which shows our most important values
at a glance. You can also compare two models with it:

```{r}
glance(gm_lm)
```

The last main fucntion from `broom` is `augment`. This gives us the original
dataset back plus the extra rows like standard errors for residuals, residuals,
etc. This function is really useful for plotting the results of the model

```{r}
augment(gm_lm) %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  # the geom line adds a line to the plot on top of the already existing plot,
  # i.e. we can give it a different x and y to plot via the aes which returns a
  # mapping
  geom_line(aes(x = gdpPercap, y = .fitted))
```

- Note here how ggplot basically crease the canvas for the plot. Then you layer
  on a point distribution using the data, then you layer on a line using the
  data and specifying the variable you want. In this case `.fitted` will give
  us all the points on the fitted line. This is the same as using ggplot's
  `geom_smooth` function. We are just doing it the long way around

## Subbing out function for e.g. `loess`

The most important takeaway here is the idea of **model flexibility**. We have
made a robust programme which let's us swap in and out functions. For example,
let's use the `loess` function to plot the local extimation line which is the
default given by ggplot's `geom_smooth`

```{r}
gm_loess <- loess(lifeExp ~ gdpPercap, data = gm_2007, span = .5)
summary(gm_loess)
```

- Again, we see the tilde notation for specifying LHS and RHS
- An interesting note. When I ran what I thought was known as the `tidy`
  **function**, `R` threw an error mentioning loess object doesn't have the
  method `tidy`. I.e. methods are defined in objects but somehow they are
  called with function notation.

::: {.example #loess name="loess example"}
<br />\hfill\break

```{r}
augment(gm_loess) %>%
  ggplot(aes(gdpPercap, lifeExp)) +
  geom_point() +
  geom_line(aes(gdpPercap, .fitted))
```

:::

---

### What is `loess` actaully doing

`loess` stands for "locally estimated scatter-plot smoothing". If we look at
the output of exmaple \@ref(exm:loess), we see that it is looking at the local
points and approximating them. `loess` has some options, for example:

- `span`: tells the function how localised to make its approximation
