---
author: "Kamal Sacranie"
numbersections: true
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: bookdown::pdf_document2
---

# Candy rankings

```{r, include=F}
library(plotly)
library(broom)
library(fivethirtyeight)
library(tidyverse)
```

It's always good practice to start by visualising some of your data. For fun
let's just fit a simple linear model to this dataset:

```{r}
cr_lm <- lm(pricepercent ~ winpercent, data = candy_rankings)
tidy(cr_lm)
```

```{r}
augment(cr_lm) %>%
  ggplot(aes(x = pricepercent, y = winpercent)) +
  geom_point() +
  geom_line(aes(y = .fitted))
```

- Nice tip with `geom` functions is that because it take in the object the
  previous function returns, you can keep the `x` value the same and reassign
  the `y`

## Adding a categorical variable to our model

```{r}
cr_lm_cat <- lm(winpercent ~ pricepercent + chocolate, candy_rankings)
tidy(cr_lm_cat)
```

- Here in the `formula` argument, we add the categorical variable `chocolate`
- This becomes a dummy variable for our regression model (`chocolateTRUE`)

Let's now represent this new categorical variable on our plot through the use
of colour:

```{r}
augment(cr_lm_cat) %>%
  ggplot(aes(
    x = pricepercent,
    y = winpercent,
    color = chocolate,
    shape = chocolate,
    linetype = chocolate
  )) +
  geom_point() +
  geom_line(aes(y = .fitted))
```

- So in the `aes` function we have a whole load of arguments which take a set
  of true or false values to boolean index the characteristics of the data

## Adding a continuous predictor and creating a 3D plot

A common package used to create 3D plots is `plotly`. 

```{r}
candy3d <- plotly::plot_ly(candy_rankings,
  x = ~pricepercent, y = ~sugarpercent,
  z = ~winpercent,
  type = "scatter3d"
)
```

> This is a perfect example of how `R` is so good out the box for data
> visualisation

We can fit our regression plane in this three dimensional space:

```{r}
cr_lm_sugar <- lm(
  winpercent ~ pricepercent + sugarpercent,
  data = candy_rankings
)

xy_plane <- expand.grid(0:100, 0:100) / 100
ps_plane <- xy_plane %>%
  rename(pricepercent = Var1, sugarpercent = Var2)

lm_plane <- augment(cr_lm_sugar, newdata = ps_plane)
lm_matrix <- matrix(lm_plane$.fitted, nrow = 101, ncol = 101)

candy3d %>%
  add_surface(
    x = ~ (0:100) / 100,
    y = ~ (0:100) / 100,
    z = ~lm_matrix
  )
```

- The `expand.grid` function creates a data frame from all combinations of the
  supplied vectors or factors
- We then rename the variable of the plane with `rename`
- We then see the `augment` function in use 
