# (APPENDIX) Appendices {-}

# (PART) Appendices {-}

# `R` basic notes

> This is the more general stuff that isn't categorisable into the lecture
> work. For specific `R` functions just look at the coursework in the root
> directory of the course

## Using the `R` repl

To use the `R` repl, you just

```{bash}
R
```

type `R` on the command line. This launches the `R` repl and you can write any
code you want. I.e. when you want to install a package, you can use the repl

## Managing packages

To install a package in `R`, you use the install object:

```{r}
install.packages({
  package:string
})

# For example installing renv for virtual environments
install.packages("renv")
```

- You pass the name of your package as the argument as a string

## `R` virtual environments

You don't want global packages everywhere, some of them will only be useful for
particular projects. Queue `renv`, the `R` virtual environment manager.

To create a virtualenv, go into the root of your project directory and run

```{r create-venv}
renv::init()
```

to initialise your virtual environment.

- You should create your `.gitignore .Rbuildignore`, and `.Rprofile` files
  before running this because it auto updates them

You can then install packages as usual and once you have completed your
project, you can use `renv::snapshow()` to write the dependencies to the
`renv.lcok` file.^[Note that this only writes the libraries you call in your
code. So without a project, you will not have any dependencies]

### LSP in virtual environments

Just spent ages trying to figure out why my LSP wasn't picking up my
virtualenv. It was because the `R` package `languageserver` must be installed
in the virtualenv in order for neovim to find the script to execute.

LSP has:

- Formatting
- Completion

## Loading with the library command

Note: in \@ref(r-virtual-environments) we preface the `init()` command with `renv::` denoting that we want
to use the `init()` command from that specific package. If we were running a
command constantly, we could use:

```{r}
library(renv)
init()
```

to **bring all the function from the `renv` package into our environment's
scope**.

## Environemnt variables

You can get and set environemnt variables with:

```{r}
Sys.getenv()
Sys.setenv()
```

To remove an envrionement variable you can use:

```{r}
Sys.unsetenv()
```

## Viewing `R` documentation

In `R`, if you want any documentation, you just need to preface the command
with a question mark:

```{r}
?install
```

- Note that we don't call the function when we ask for help

## The `select` function

You can use the `select` function to select/deselect rows from your model:

```{r}
subset <- select(my_data, -"name of column")
```

## Indexing in `R`

Indexing in `R` is pretty similar to something like `pandas`. If I have a
dataset I can index it in one of two way. The first:

```{r}
myplot[5, ]
myplot[, 5]
myplot[2, 5]
myplot[2:4, 5:7]
myplot[myplot[, 5] == 7, 5]
```

- Line 1 returns row 5 and all columns
- Line 2 returns column 5 with all rows
- Line 3 returns cell from column 5 row 2
- Line 4 returns the cells in rows 2 to 4 and columns 5 to 7^[Not sure if it's
  inclusive or not]
- Line 5 return the rows in column 5 which have a value of 7

> Data indexing in `R` is always [<rows>, <cols>]

The second way is with the `$` operator:

```{r}
mydata[mydata$year == 2020, ]
```

- This will return all columns for the rows that contain 2020

### The `tidyverse` way of indexing

This is quite a javascript way of doing things. We use the `filter` function,
much like you would use it in js

```{r}
filter(dataset, c(year == 2000, continent == "Asia", ...))
```

## Matrices and vectors

In `R` you can create a matrix using the `matrix` function:

```{r, eval=T}
x <- matrix(data = 1:9, nrow = 3, ncol = 3)
matrix(1:9, nrow = 3, ncol = 3, byrow = T)
```

- We see here that we specify the data as a vector^[This is kind of like a list
  but it only holds one type of data type and is strictly one dimentional] of
  numbers
- We also specify the number of rows and columns and the way in which the data
  is spread across the columns/rows


### Matrix multiplication

In `R`, you can multiply matrices together with the `%*%` operator. We can
check the dimensions of our matrix:

```{r, eval=T}
dim(x)
```

> **NB**: your matrices MUST be able to be mathematically multipliable of
> course (cols `==` rows). You must be careful when multiplying because `R`
> will not throw an error message if one of the dimentions are a multiple of
> the other

One of the **frustrating things** about `R` is that if you use the `dim`
function on a list object, it will say that it has no dimensions, but if you
use it in a function that treated it as a vector^[Vectors have to have defined
dimensions], it will have $0\times 0$ dimensoins:

```{r, eval=T}
dim(rep(1, 3))
x %*% rep(1, 3) # The list auto converts to a vector here
```

- `rep` is how you define a list of the same number repeating themselves for
  the number of times you pass in as the second argument

### Transpose and symmetry

Recall: even if a matrix $\mathbf A$ is not square, both $\mathbf A^{T}\mathbf
A$ and $\mathbf A\mathbf A^{T}$ **are square and symmetric**^[A matrix
multiplied by it's transpose is always square]:

```{r, eval=T}
A <- matrix(c(rep(1, 4), 0, 0), nrow = 3)
t(A) %*% A
```

- This matrix is invertable but `A %*% t(A)` is not

### Pseudoinversion

```{r, eval=T}
MASS::ginv(A)
```

- This is a generalised inverted matrix which can act like a "one sided"
  inverse on matrix `A`

The generalised inverse only works when multiplying on the left to give the
$2\times 2$ identity matrix:

```{r, eval=T}
ginv(A) %*% A
```

## Miscellaneous tips

### Viewing your dataset in a spreadsheet-like manner

```{r}
View(<dataset>)
```

### Generating a polynomial

You can generate a polynomical using:

```{r}
poly({datacolumn}, {polynomial degree})
```

- This will basically plot a standard polynomial for the `x` data values
  (vector?) that you pass as an arg.

### Adding cols onto a dataframe

You can use `augment` to add columns onto a dataframe

```{r}
augment(my_df, newdata = newdf)
```

### Including all variables in your model

The shorthand to inlcude all the variables in your dataset in your model is:

```{r}
lm(y ~ ., data = my_data)
```

What you need to avoid here though is using a variable which is unique to every
piece of data, like brand, etc.

### `rbind`

Take a sequence of vector, matrix or data-frame arguments and combine by
columns or rows, respectively. We are basically concatenating
tables/matrices/vectors.

```{r}
rbind(<object-1>, <object-2>)
```
