# `R` notes^[These are probably temporary and will go into own book]

> This is the more general stuff that isn't categorisable into the lecture
> work. For specific `R` functions just look at the coursework in the root
> directory of the course

## Using the `R` repl

To use the `R` repl, you just

```{bash}
R
```

type `R` on the command line. This launches the `R` repl and you can write any
code you want. I.e. when you want to install a package, you can use the repl

## Managing packages

To install a package in `R`, you use the install object:

```{r}
install.packages({
  package:string
})

# For example installing renv for virtual environments
install.packages("renv")
```

- You pass the name of your package as the argument as a string

## `R` virtual environments

You don't want global packages everywhere, some of them will only be useful for
particular projects. Queue `renv`, the `R` virtual environment manager.

To create a virtualenv, go into the root of your project directory and run

```{r create-venv}
renv::init()
```

to initialise your virtual environment.

- You should create your `.gitignore .Rbuildignore`, and `.Rprofile` files
  before running this because it auto updates them

You can then install packages as usual and once you have completed your
project, you can use `renv::snapshow()` to write the dependencies to the
`renv.lcok` file.^[Note that this only writes the libraries you call in your
code. So without a project, you will not have any dependencies]

### LSP in virtual environments

Just spent ages trying to figure out why my LSP wasn't picking up my
virtualenv. It was because the `R` package `languageserver` must be installed
in the virtualenv in order for neovim to find the script to execute.

LSP has:

- Formatting
- Completion

## Loading with the library command

Note: in \@ref(r-virtual-environments) we preface the `init()` command with `renv::` denoting that we want
to use the `init()` command from that specific package. If we were running a
command constantly, we could use:

```{r}
library(renv)
init()
```

to **bring all the function from the `renv` package into our environment's
scope**.

## Environemnt variables

You can get and set environemnt variables with:

```{r}
Sys.getenv()
Sys.setenv()
```

To remove an envrionement variable you can use:

```{r}
Sys.unsetenv()
```

## Viewing `R` documentation

In `R`, if you want any documentation, you just need to preface the command
with a question mark:

```{r}
?install
```

- Note that we don't call the function when we ask for help

## The `select` function

You can use the `select` function to select/deselect rows from your model:

```{r}
subset <- select(my_data, -"name of column")
```

## Miscillaneous tips

### Viewing your dataset in a spreadsheet-like manner

```{r}
View(<dataset>)
```

### Generating a polynomial

You can generate a polynomical using:

```{r}
poly()??????
```

### Adding cols onto a dataframe

You can use `augment` to add columns onto a dataframe

```{r}
augment(my_df, newdata = newdf)
```

### Including all variables in your model

The shorthand to inlcude all the variables in your dataset in your model is:

```{r}
lm(y ~ ., data = my_data)
```

What you need to avoid here though is using a variable which is unique to every
piece of data, like brand, etc.
